Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-01 analysis/paper/paper.Rmd
work on introduction

- the definition of craft specialization

The emergence of craft specialization is usually associated with the changes in a socio-economic system of a society since it links to the organization of production

context, concentration, scale, intensity

Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.

Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism. One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp. 334; @Stark1995; @Costin1991].

social changes -  culture contact framework 
- the cause of emergence of craft specialization, elite control, rice intensification, political, economic 
- here I have the third cultural resistance, identity take a look at social boundary
- standardization as indicator, metric and technological
- GM is more sensitive for subtle changes, and the outline
- innovative: CV test 
- hypothesis, goals

Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized. Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism. One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp. 334; @Stark1995; @Costin1991].

Several variables have been proposed to measure ceramic standardization, including metric, compositional, and technological attributes [@Rice1991; @Blackman1993; @Costin1991; @Hirshman2010; @Tite1999]. However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization. Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan. We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.

@@

The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornament elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007].

More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. There are only two forms of vessels, one is cooking pot and another one is steamer made of two pots stacking together. The pots show high consistency in shape - globular body with short neck and wide mouth. A wide variety of impressed geometric motifs are applied all over the exterior body surface. They were believed used for cooking according to the evidence of charred residue commonly observed on interior surface and soot on exterior surface. Besides, there is no other utilitarian vessels found at this site. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Those pots might be pinched and shaped by hand according to the pressed finger marks usually observed inside of pots.

>>>>

The development of craft specialization is usually associated with the social changes since it reflects production systems that inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. According to the political approaches, craft specialization, attached production specifically, is viewed as a strategy used by elites to control over production systems for the sake of power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. Economic models consider production, exchange, and consumption as a whole and explain that specialization is a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], or a reaction to the changes in subsistence practices [@Stark1995economi]. However, the explanation for the emergence of specialized production emerges might not be exclusive or straightforward. Also, specialization is not necessarily accompanied with social differentiation that is observed in lowland northern Philippines with low degree of ceramic specialization during the Spanish colonial period [@Acabado2018]. This implies that the condition and contexts under which specialists organized are important for interpretation of changes in social organization [@Roux2018].

- standardization as indicator, metric and technological
- GM is more sensitive for subtle changes, and the outline
- innovative: CV test

craft specialization is commonly identified through statistical standardization of specific dimensions of a particular item – standardization as evidence of specialization.

The variation of standardization of specialized product is commonly used as indicator to identify craft specialization based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Arnold2000, pp. 334; @Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, including metric, compositional, and technological approaches [@Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, linear measurements is widely used by calculation coefficient of variation (CV) that lower CV value indicates higher degree of standardization [@Hirshman2010; @Eerkens2001]. However, traditional typological and linear measurements are limited because they can be insensitive to subtle variations resulting from changes in craft specialization.

Taking a new approach to the measurement of craft specialization, we studied shape and standardization of locally made pottery to identify changes in pottery production at Kiwulan (1400-1900 AD), a large multi-component archaeological site in northeastern Taiwan. We apply geometric morphometric methods to study artifact shapes to investigate if there are any changes resulting from foreign contact of European and Chinese that might indicate social changes in the indigenous society.

- social changes -  culture contact framework 
- here I have the third cultural resistance, identity take a look at social boundary
- hypothesis, goals

This paper

@@

The upper component experienced frequent foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. The excavation revealed abundant artifacts in which potsherds are the dominate materials throughout the site. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were also commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features were also found, such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007]. The predominant artefact found at Kiwulan is locally made ceramics, which distributed throughout the sequence across the site.

More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter in between. Those pots show high consistency in shape - globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking because of the evidence of charred residues or carbon deposits commonly observed on the interior and soot on the exterior. Besides, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.
